\section{Program Safety via Execution Trees}
\label{sec:program-safety}

In this section,
we present a theoretical foundation that enables us to prove safety of programs based on properties of the symbolic semantics.
The main result of this section is the \emph{program safety} theorem (\Cref{theorem:safety-via-et}),
which states that a program is safe with respect to the concrete semantics if the execution tree of the program satisfies certian safety properties.

\subsection{Module Assumptions}
\label{sec:module-assumptions}

To simplify the reasoning over the semantics of \llvm IR,
we introduce additional assumptions about the structure of the \llvm module,
as formally defined below.\footnote{
Additional assumptions about \textit{struct} offsets in $\tgep$ instructions are encoded in our \coq framework.
}

\begin{definition}
\label{def:supported}
An \llvm expression is \emph{supported} if it is not a $\tpoison$ value and does not contain potentially erroneous operations ($\opcode{udiv}$, $\opcode{shl}$, $\tgep$, \etc).
A module $\module$ is \emph{supported} if $\module$ does not contain $\tundef$ values and every instruction operand in $\module$ is a supported \llvm expression.\footnote{
For example, $\codequote{r ~ = ~ \opcode{udiv} ~ x, ~ y}$ is allowed, but $\codequote{r ~ = ~\opcode{add} ~ (\opcode{udiv} ~ x, ~ y), ~ z}$ is disallowed.
}
%\begin{enumerate}
%    \item Every operand of an instruction in $\module$ is a \emph{supported} \llvm expression.
%    \item There are no $\tundef$ values in $\module$.
%\end{enumerate}
\end{definition}

%In the rest of this paper, we assume that the module $\module$ is \emph{supported}.
In practice, the instruction operands in \llvm IR programs generated by the compiler are either variables or constants.
Furthermore, the SE engine replaces $\tundef$ values with pre-defined constants before starting the analysis,
which is allowed by the \llvm reference.
Therefore, the assumptions described in~\Cref{def:supported} are well-justified for parctical applications.

%In addition,
%if the indexed type of a given index in a $\tgep$ instruction is a struct type,
%then that index must be a constant and lie within the range of $\bvsort{64}$.
%As a matter of notation, we say that a module is \emph{supported} if it satisfies all these assumptions.

%If the indexed type of a given index in a $\tgep$ instruction is a struct type,
%then that index must be a constant and lie within the range of $\bvsort{64}$.
%At a high level, this assumption is required in order to handle the case of $\tgep$ in the proof of the reachability completeness theorem.

\subsection{Reachability Completeness Theorem}
\label{sec:completeness}

In this section,
we establish the \emph{reachability completeness} of the symbolic semantics with respect to the concrete semantics.
More specifically, our main result is~\Cref{theorem:completeness},
which states that any sequence of derivations in the concrete semantics that does not introduce $\tpoison$ values can be simulated by a corresponding sequence of derivations in the symbolic semantics.
Ultimately, this correspondence will help us proving the main theorem of~\Cref{sec:program-safety} (\Cref{theorem:safety-via-et}).

To establish reachability completeness,
we first define a relation between concrete and symbolic states.
Informally, we say that a symbolic state $s$ \emph{represents} a concrete state $c$ if $c$ can be obtained from $s$ via concretization using a model that satisfies the path constraints of $s$.
The formal definition is given below.

% TODO: same number of memory objects
\begin{definition}
\label{def:over-approx-state}
Let $m$ be a model.
An SMT expression $e$ \emph{represents} a value $v \in \Vals$ via $m$,
denoted by $e \overapproxvia{m} v$, if $\evalmodel{m}{e} = v$.
A symbolic memory object $(\addr, \size, \smtarray)$ represents a memory object $(\addr, \size, \arrayvalue)$ via $m$ if
$\arrayselect{\smtarray}{\bvconst{\indexwidth}{k}}$ represents $\arrayvalue(\valoftype{k}{\indextype})$ via $m$ for every $k$ in the range of the index type.
A symbolic heap $\symmemory$ represents a heap $\memory$ via $m$, denoted by $\symmemory \overapproxvia{m} \memory$,
if the symbolic memory objects of $\symmemory$ represent the memory objects of $\memory$ via $m$.
A symbolic state $s \triangleq \symstate{\location}{\symlocalstore}{\symstack}{\symmemory}{\pc}$ \emph{represents} a concrete state $c \triangleq \progstate{\location}{\localstore}{\stack}{\memory}$ via $m$,
denoted by $s \overapproxvia{m} c$,
if $\symlocalstore \overapproxvia{m} \localstore$,
$\symmemory \overapproxvia{m} \memory$,
and $\evalmodel{m}{\pc} = \true$.
Furthermore, $s$ \emph{represents} $c$, denoted by $s \overapprox c$,
if $s \overapproxvia{m} c$ for some model $m$.
\end{definition}

Based on~\Cref{def:over-approx-state},
one can see that a concrete state containing a $\tpoison$ value cannot be represented by a symbolic state.
This implies that we cannot establish the reachability completeness of the symbolic semantics with respect to the concrete semantics.
%Therefore, a sequence of derivations in the concrete semantics involving $\tpoison$ values cannot be simulated by a sequence of derivations in the symbolic semantics.
Instead, we establish the reachability completeness with respect to a subset of the concrete semantics restricted to concrete states that do not contain $\tpoison$ values.
%Formally, it is defined as follows:

In the following,
we formally define the restricted concrete semantics.
We first define \emph{poison-free} states, and then define the corresponding step relation.

\begin{definition}
A concrete state $c \triangleq \progstate{\location}{\localstore}{\stack}{\memory}$ is \emph{poison-free}
if $\localstore$ and $\memory$ do not contain $\tpoison$ values.
\end{definition}

\begin{definition}
\label{def:ns-step}
We say that there is a \emph{poison-free} step between two concrete states $c$ and $c'$, denoted by $c \nsstep c'$,
if $c \step c'$ and both $c$ and $c'$ are poison-free.
\end{definition}

Similarly to~\Cref{def:safety},
we define when a program, \ie an \llvm module, is considered safe with respect to the restricted concrete semantics.
The formal definition is given below.

\begin{definition}
\label{def:issafe-ns-step}
Let $c_0 \triangleq \initstate(\module)$,
then $\module$ is \emph{safe} with respect to the restricted concrete semantics, denoted by $\issafens(\module)$, if:
\[ \forall c. \ c_{0} \multinsstep c \implies \neg \errorstate(c) \]
(The relation $\multinsstep$ is the reflexive-transitive closure of the relation $\nsstep$.)
\end{definition}

In general, some derivations are disallowed by the restricted concrete semantics but allowed by the full concrete semantics.
However, under certain conditions, the restricted and full concrete semantics align.
This is formulated in the following lemma.

\begin{lemma}
\label{lemma:multi-ns-step-relative-completeness}
Let $c$ and $c'$ be concrete states.
If $\module$ is supported, $c$ is poison-free, no error states can be derived from $c$ using the $\multinsstep$ relation, and $c \multistep c'$,
then $c \multinsstep c'$.
\end{lemma}

We first establish single-step reachability completeness.
To do so, we must assume that the concrete states are \emph{well-formed}.
Intuitively, this means that the address intervals in the heap do not overflow and do not intersect.
The formal definition is given below.

\begin{definition}
\label{def:well-formed}
A memory object $(\addr, \size, \arrayvalue)$ is \emph{well-formed} if its address interval does not wrap around the address space,
\ie $\addr + \zeroext{\bvsort{64}}{\size}$ does not overflow.
A heap $\memory$ is \emph{well-formed} if all its memory objects are well-formed and there are no memory objects with overlapping address intervals.
A state $\progstate{\location}{\localstore}{\stack}{\memory}$ is well-formed if $\memory$ is well-formed.
\end{definition}

Now, with the help of~\Cref{def:well-formed},
we formally define the lemma that establishes single-step reachability completeness.

\begin{lemma}
\label{lemma:completeness-step}
Let $c$ be a concrete state and $s$ be a symbolic state.
If $c$ is well-formed, $s \overapprox c$ and $c \nsstep c'$,
then there exists a symbolic state $s'$ such that $s \symstep s'$ and $s' \overapprox c'$.
\end{lemma}

Finally, we extend~\Cref{lemma:completeness-step} to establish reachability completeness for arbitrary sequences of derivations starting from the initial state.
The formal definition is given below.

\begin{theorem}
\label{theorem:completeness}
Let $c_{0} \triangleq \initstate(\module)$, and let $s_{0} \triangleq \initsymstate(\module)$.
If $\module$ is supported and $c_{0} \multinsstep c$,
then there exists a symbolic state $s$ such that $s_{0} \symmultistep s$ and $s \overapprox c$.
\end{theorem}

As we will see next,
\Cref{lemma:multi-ns-step-relative-completeness} and \Cref{theorem:completeness} will helps us establish the \emph{program safety} theorem.

\subsection{Program Safety Theorem}
In this section,
we establish the \emph{program safety} theorem,
which states that a program is safe with respect to the concrete semantics if the execution tree of the program satisfies certian safety properties.

The formal definition of \emph{execution trees} is given below.
\begin{definition}
The \emph{execution tree} of a program records the paths explored by the \SE engine during the analysis of that program,
where each node corresponds to a symbolic state computed by the \SE engine.
An execution tree $t$ with a root symbolic state $t.\attrstate$ and a list of successor subtrees $t.\attrchildren$ is denoted by $\tree(s, l)$,
where $s$ and $l$ correspond to the respective attributes.
\end{definition}

Our initial definition of the safety properties for execution trees is given below.
As we will see later, this definition will turn out to be too restrictive for practical use,
and we will refine it in~\Cref{def:safe-et}.

\begin{definition}
\label{def:safe-et-strict}
Let $s$ be a symbolic state and $l$ a list of execution trees.
We say that $tree(s, l)$ is \emph{safe}, denoted by $\safeetstrict(\tree(s, l))$ if:
\begin{enumerate}
    \item $s$ is not an error state
    \item if $s \symstep s'$ then either the path constraints of $s'$ are unsatisfiable or there exists a subtree $t \in l$ such that $\safeet(t)$ and $s' = t.\attrstate$.
\end{enumerate}
\end{definition}

Based on this definition,
we establish a weaker version of the \emph{program safety} theorem,
which guarantees safety with respect to the restricted concrete semantics.
The formal definition is given below.

\begin{lemma}
\label{lemma:safety-via-et-strict-ns}
Let $\module$ be a supported module, let $s_0 \triangleq \initsymstate(\module)$, and let $l$ be a list of execution trees.
If $\safeetstrict(\tree(s_0, l))$, then $\issafens(\module)$.
\end{lemma}

To see intuitivily why this lemma holds, consider the single-step case.
Let $s$ be a symbolic state and $c$ be a concrete state such that $s \overapprox c$,
and suppose that the subtree $\tree(s, l)$ is safe.
If $c \nsstep c'$, then from~\Cref{theorem:completeness},
it follows that there exists $s'$ such that $s \symstep s'$ and $s' \overapprox c'$.
From~\Cref{def:safe-et-strict}, it follows that $s'$ is not an error state,
and since $s' \overapprox c'$, then $c'$ is not an error state.

Now, let $c_0 \triangleq \initstate(\module)$ be the initial concrete state.
If $\issafens(\module)$, then no error states can be derived from $c_0$.
Next, if $c_0 \step c$,
then based on~\Cref{lemma:multi-ns-step-relative-completeness}, it follows that $c_0 \nsstep c$.
Finally, we know that $c$ is not an error state, so, $\issafe(\module)$.
This argument allows us to establish a stronger version of~\Cref{lemma:safety-via-et-strict-ns},
which is given below.

\begin{lemma}
\label{lemma:safety-via-et-strict}
Let $\module$ be a supported module, let $s_0 \triangleq \initsymstate(\module)$, and let $l$ be a list of execution trees.
If $\safeetstrict(\tree(s_0, l))$, then $\issafe(\module)$.
\end{lemma}

However, the formulation of~\Cref{lemma:safety-via-et-strict} is not practical enough.
In general, the symbolic states computed by the \SE engine may differ from those derived by the mechanized symbolic semantics,
and in that case, the precondition of~\Cref{lemma:safety-via-et-strict} will not hold.
The address this,
we refine~\Cref{def:safe-et-strict} by requiring \emph{equivalnece} instead of \emph{equality}.

In the following, we formally define equivalence between symbolic states.

\begin{definition}
The symbolic stores $\symlocalstore_1$ and $\symlocalstore_2$ are \emph{equivalent}, denoted by $\symlocalstore_1\equiv \symlocalstore_2$, if they are defined over the same variables,
and for each such variable $x$, $\mapaccess{\symlocalstore_1}{x} \equiv \mapaccess{\symlocalstore_2}{x}$.
The memory objects $(\addr, \size, \smtarray_1)$ and $(\addr, \size, \smtarray_2)$ are equivalent if $\smtarray_1 \equiv \smtarray_2$.
The heaps $\symmemory_1$ and $\symmemory_2$ are equivalent, denoted by $\symmemory_1 \equiv \symmemory_2$,
if their memory objects are equivalent.
The states $s_1 \triangleq \symstate{\location_1}{\symlocalstore_1}{\symstack_1}{\symmemory_1}{\pc_1}$ and $s_2 \triangleq \symstate{\location_2}{\symlocalstore_2}{\symstack_2}{\symmemory_2}{\pc_2}$ are equivalent,
denoted by $s_1 \equiv s_2$, if $\location_1 = \location_2$, $\symlocalstore_1 \equiv \symlocalstore_2$, $\symmemory_1 \equiv \symmemory_2$, and $\pc_1 \equiv \pc_2$.
\end{definition}

Now, the refined version of the safety properties of execution trees is defined below.

\begin{definition}
\label{def:safe-et}
Let $s$ be a symbolic state and $l$ a list of execution trees.
We say that $tree(s, l)$ is \emph{safe}, denoted by $\safeet(\tree(s, l))$ if:
\begin{enumerate}
    \item $s$ is not an error state
    \item if $s \symstep s'$ then either the path constraints of $s'$ are unsatisfiable or there exists a subtree $t \in l$ such that $\safeet(t)$ and $s' \equivstate t.\attrstate$.
\end{enumerate}
\end{definition}

%This property ensures that there are no error states in the tree, and furthermore,
%that every feasible successor $s'$ of any state $s$ occurring in the tree is also represented in the tree as a successor of $s$.
%Infeasible states can be ignored, as they do not represent any concrete states.

%The following theorem states that if the execution tree of the program satisfies the safety property described above,
%then the program is safe w.r.t. the concrete semantics.

Based on this refined definition,
we can apply the reasoning from~\Cref{lemma:safety-via-et-strict} to establish the desired \emph{program safety} theorem.

\begin{theorem}
\label{theorem:safety-via-et}
Let $\module$ be a supported module, let $s_0 \triangleq \initsymstate(\module)$, and let $l$ be a list of execution trees.
If $\safeet(\tree(s_0, l))$, then $\issafe(\module)$.
\end{theorem}
