\section{Discussion}
\label{sec:discussion}

In this section,
we discuss the key aspects and implications of our work.

\paragraph{Applicability to other \SE engines}
% our work provide the tools to address semantic gaps
% for performance, depends on the gap size
While our current implementation targets \klee, %our underlying architecture is modular.
we believe that our approach can be adapted to other \SE engines whose semantics differ from those of \klee.
As demonstrated in the case of \klee, our approach is designed to bridge such semantic gaps.
%Indeed, \klee's symbolic semantics differs significantly from our symbolic semantics (\Cref{sec:symbolic-semantics}),
%but our approach is designed to bridge this gap.
First, we develop a theory that allows us to prove that the SMT transformations in \klee preserve semantic equivalence.
Second, we use interface lemmas which help to bridge the gap between \klee's optimized memory modeling and our mechanized memory modeling.
In general, the gap between the symbolic semantics of a given \SE engine and our mechanized symbolic semantics has an impact on performance,
as it affects both proof generation and proof compilation.
In \SE engines where this gap is substantial, the performance might decrease.
Otherwise, the performance is expected to be comparable to our results.

\paragraph{Formalized Theory}

%The theoretical contributions of our formalized theory extend beyond our specific application in the context of proof logging.
Beyond its application to proof logging, our formalized theory (\Cref{sec:concrete-semantics,sec:symbolic-semantics,sec:program-safety}) stands as an independent theoretical contribution.
%Our formalized theory has theoretical contributions that extend beyond its specific application to proof logging.
First, the concrete semantics can serve as a foundation for proving program safety using other techniques,
such as deductive verification or abstract interpretation.
Second, the symbolic semantics can provide a robust framework for formalizing advanced extensions of symbolic execution,
including state merging, alternative memory models, and alternative memory allocation schemes.

% TODO: modeling of malloc
