\section{Implementation}
\label{sec:implementation}

% TODO: we added a new allocator to be aligned with...

Our implementation consists of two main components:
a framework in \coq (9.0.1) and an extension of the \SE engine.
Our framework is based on~\cite{klee-rocq},
and it includes the definitions of the semantics (\Cref{sec:concrete-semantics,sec:symbolic-semantics}) and the theoretical foundation (\Cref{sec:program-safety,sec:proof-generation}).
Our proof generation approach (\Cref{sec:proof-generation}) is implemented on top of \klee~\cite{klee},
a \emph{state-of-the-art} \SE engine operating on \llvm IR~\cite{llvm}.
Currently, some of the SMT transformations used in \klee are not supported by our implementation,
but we plan to support them in future versions.
In the $\tgep$ instruction, our implementation currently supports only the \opcode{nuw} attribute.

\subsection{\coq Framework}
Our \coq framework consists of approximately 40,000 lines of code (LOC),
excluding third-party dependencis (\eg CompCert~\cite{CompCert} integers).
For comparison,
this is more than twice the size of the framework of~\cite{klee-rocq}, which consists of approximately 18,000 LOC.
This substantial difference reflects the additional complexity of our framework.
To the best of our knowledge,
our \coq framework is one of the largest \coq formalizations in programming language theory (\eg CompCert~\cite{CompCert}, VST~\cite{vst}, Iris~\cite{irisproject}, and $CH_2O$~\cite{krebbers2015typed}).
% main theorem: 3.5K
% completeness: 6.2K
% simplification theory: 8.1K
% automation: 6.7K
% non-spec: 1.9K
% module assumptions: 0.5K
% well-formedness: 0.9K
To provide a detailed insight into the core components of our framework,
we report the size (LOC) of those components:
The main theorem (\Cref{theorem:safety-via-et}),
the completeness theorem (\Cref{theorem:completeness}),
the SMT simplification theory (\Cref{sec:proof-et}),
and the interface lemmas (\Cref{sec:proof-et}) approximately consist of 3,500 LOC, 6,200 LOC, 8,100 LOC, and 6,700 LOC, respectively.
%\begin{itemize}
%    \item Main theorem (\Cref{theorem:safety-via-et}): $\approx$3,500 LOC
%    \item Completeness theorem (\Cref{theorem:completeness}): $\approx$6,200 LOC
%    \item SMT simplification theory (\Cref{sec:proof-et}): $\approx$8,100 LOC
%    \item Interface lemmas (\Cref{sec:proof-et}): $\approx$6,700 LOC
%\end{itemize}

\subsection{Implementation Issues in \klee}
Thanks to our formalization efforts,
we found several implementation issues in \klee that relate to memory modeling, which we discuss next.

\textit{Issue 1: Inconsistent Modeling of the Index Type.}
On one hand,
\klee uses SMT arrays of sort $\arraysort{\smtsort{32}}{\smtsort{8}}$ to represent the contents of memory objects,
which suggests that the modeled index type is $\bvsort{32}$.
On the other hand,
\klee computes the offsets in the interpretation of $\tgep$ using SMT expressions of sort $\smtsort{64}$,
which suggests that the modeled index type is $\bvsort{64}$.

\textit{Issue 2: Incorrect Interpretation of \textup{$\tgep$}}
\klee does not account for overflows in the offset computation, thus not fully compliant with the \llvm reference.
To fix this,
we modified the offset computation to conform to the index type $\bvsort{32}$,
ensuring consistency with the index sort $\smtsort{32}$ used in the SMT arrays,
and we report an error when such overflows occur.

\textit{Issue 3: Incorrect Interpretation of Memory Allocations}
According to the \llvm semantics,
if the allocation size exceeds the largest \emph{signed integer} representable by the index type,
then the result is undefined behavior.
However, \klee currently does not check this condition.

\textit{Issue 4: Silent Truncation in Memory Allocations}
\klee silently truncates the allocation size to the \code{uint32\_t} type,
without reporting any error to the user, which may lead to unexpected behavior.
For example, if we allocate an object of size $2^{32} + 10$,
then the allocation size is internally cast to $10$,
and if later we access that object at offset $10$, then we will get an out-of-bounds error, which is a false positive.

%To fix this, we modified the interpretation of $\tgep$ as follows:
%First, we modified the offset computation to conform to the index type $\bvsort{32}$,
%ensuring consistency with the index sort $\smtsort{32}$ used in the SMT arrays.
%Second, if the \code{nuw} attribute is enabled, we report an error when an overflow occurs.

%The lack of overflow checks has some additional consequences,
%as can be demonstrated by~\Cref{figure:klee-issue}.
%After executing the $\tgep$ instruction at~\cref{line:issue-llvm-gep}, the value of \code{p2} is $a + 4294967296$,
%assuming that the value of \code{p1} is $a$.
%Since \klee ignores the overflow in this case, it continues to the next instruction at~\cref{line:issue-llvm-store}.
%When \klee accesses a memory object (\ie $\tload$ and $\tstore$),
%the SMT expression of the offset is always truncated to an SMT expression of sort $\smtsort{32}$.
%So in this case,
%the SMT expression $4294967296$ (of sort $\smtsort{64}$) is truncated to the SMT expression $0$ (of sort $\smtsort{32}$).
%Therefore, this $\tstore$ instruction actually updates the first byte of the memory object allocated at~\cref{line:issue-llvm-malloc}.
%We note that this behavior occurs in \klee without reporting any errors to the user.

%\begin{figure}[t]
%\begin{minipage}[t]{0.44\columnwidth}
%\lstinputlisting[basicstyle=\scriptsize\ttfamily,language=c,mathescape=true,breaklines=false]{code/alloc.simple.c}
%\end{minipage}
%\begin{minipage}[t]{0.52\columnwidth}
%\lstinputlisting[basicstyle=\scriptsize\ttfamily,language=c,mathescape=true,breaklines=false]{code/alloc.simple.ll}
%\end{minipage}
%\caption{A simple C program and its translation to \llvm IR.}
%\label{figure:klee-issue}
%\end{figure}

%% TODO: fix...
%\begin{algorithm}[t]
%\caption{The algorithm for handling $\tgep$ in \klee.}
%\label{alg:gep-klee}
%\begin{algorithmic}[1]
%
%\Function{compute-offset}{$\tau, o, \overline{j}$}
%    \If {$|\overline{j}| = 0$}
%        \State \Return $o$
%    \EndIf
%    \State $\overline{j} \gets j_1 :: \overline{j'}$
%    \If {$\tau = \arraytype{\tau'}{n}$}
%        \State $o \gets \bvadd_{64}(o, \bvmul_{64}(\bvsext_{64}(j_1), \bvconst{64}{|\tau'|}))$ \label{line:array-index}
%        \State \Return \Call{compute-offset}{$\tau', o, \overline{j'}$}
%    \EndIf
%    \If {$\tau = \structtype{\tau}{n}$}
%        \State ...
%    \EndIf
%    \State ...
%\EndFunction
%
%\Function{compute-pointer}{$\tau, p, j_0, \overline{j}$}
%    \State $o \gets \bvmul_{64}(\bvsext_{64}(j_0), \bvconst{64}{|\tau|})$ \label{line:first-index}
%    \State $r \gets \Call{compute-offset}{\tau, o, \overline{j}}$
%    \State \Return $\bvadd_{64}(p, r)$
%\EndFunction
%
%\end{algorithmic}
%\end{algorithm}
