\section{Symbolic Semantics}
\label{sec:symbolic-semantics}

In this section,
we define our symbolic operational small-step semantics of \llvm IR based on the syntax in~\Cref{figure:llvm-syntax}.

\subsection{Symbolic Memory Modeling}
\label{sec:symbolic-memory-modeling}

First, we define the major components of our symbolic memory model.
At a high level, this model differs from its concrete counterpart (\Cref{sec:memory-modeling}) by using SMT expressions instead of concrete values (\ie bit-vectors).

\begin{definition}
A symbolic memory object is defined as a tuple:
\[ (\addr, \size, \smtarray) \in \SymMemoryObjects \quad \quad \SymMemoryObjects \triangleq \pointertype \times \indextype \times \arraysort{\smtsort{\indexwidth}}{\smtsort{8}} \]
which resembles the definition of a concrete memory object,
but now the contents are encoded using an SMT array whose values are \emph{symbolic bytes} (\ie SMT expressions of sort $\smtsort{8}$).
A symbolic heap $\symmemory \in \SymHeaps$ is defined as a set of symbolic memory objects.
\end{definition}

Now, we define the operations provided by the symbolic memory model:
allocation, deallocation, read, and write.

\paragraph{Allocation}
The \emph{allocation} operation is defined using the following function:
\[ \symallocate : \SymHeaps \times \bvsort{\indexwidth} \to \SymHeaps \times \bvsort{\pointerwidth} \]
The function $\symallocate$ receives a symbolic heap $\symmemory$ and an allocation size $\size$,
and attempts to provide a symbolic memory object $\mobj{\addr}{\size}{\smtarray}$ that satisfies the following:
\begin{enumerate}
    \item $[\addr, \addr + \size)$ is fully contained within the address space (\ie $\addr < \addr + \size$)
    \item $[\addr, \addr + \size)$ does not overlap with address intervals of existing memory objects
    \item $s \geq 0$ and does not exceed the largest signed integer representable by the index type
\end{enumerate}
If such a symbolic memory object can be provided,
then the result is the tuple $\langle \symmemory \cup \mobj{\addr}{\size}{\smtarray}, \addr \rangle$.
Otherwise, the result is $\none$.
The SMT array $\smtarray$ is set to the constant array $\arrayconst{\arrayinitvalue}$.

\paragraph{Deallocation}
The \emph{deallocation} operation is defined using the following function:
\[ \symdeallocate : \SymHeaps \times \SymMemoryObjects \to \SymHeaps \]
The function $\symdeallocate$ receives a heap $\symmemory$ and a memory object $\symmo$,
and returns $\symmemory \setminus \{\symmo\}$.

\paragraph{Read}
The \emph{read} operation is defined using the following function ($w$ is the bit-width of $\tau$):\footnote{
As in~\Cref{sec:memory-modeling}, we assume that $\tau$ is one of the built-in integer types (\bvsort{8}, \bvsort{16}, \bvsort{32}, \bvsort{64}).
}
\[ \symmoreadfunc{\tau} : \SymMemoryObjects \times \smtsort{\pointerwidth} \to \smtsort{w} \]
The $\symmoreadfunc{\tau}$ function receives a memory object $\symmo$ and a pointer $p$.
First, it reads the consecutive $|\tau|$ cells starting from $p$ into a sequence of symbolic bytes.
Then, it translates them into a \textit{concat} expression of the appropriate sort.

\paragraph{Write}
The \emph{write} operation is defined using the following function:
\[ \symmowritefunc{\tau} : \SymMemoryObjects \times \pointertype \times \smtsort{w} \to \SymMemoryObjects \]
The $\symmowritefunc{\tau}$ function receives a memory object $\symmo$ (\ie $\mobj{\addr}{\size}{\smtarray}$), a pointer $p$, and an SMT expression $e$.
First, it serializes $e$ into a sequence of symbolic bytes.
Then, it uses these symbolic bytes to create a new symbolic memory object $\mobj{\addr}{\size}{\smtarray'}$,
where $\smtarray'$ is obtained by extending $\smtarray$ with additional store expressions at the corresponding offsets.

\paragraph{Memory Allocation Scheme}
The symbolic memory allocator operates linearly, mirroring the memory allocator defined in~\Cref{def:memory}.

\subsection{Symbolic Operational Small-Step Semantics}

In the following,
we define the symbolic state and the derivation rules of the symbolic operational small-step semantics.

\begin{definition}
\label{def:symbolic-state}
A symbolic state is a tuple $\symstate{\location}{\symlocalstore}{\symstack}{\symmemory}{\pc}$.
The location $\location \in \Locs$ is defined as before.
The concrete store $\localstore \in \Stores$ is replaced with a symbolic store $\symlocalstore \in \SymStores$ that maps variables to SMT expressions.
The concrete heap $\memory \in \Heaps$ is replaced with a symbolic one $\symmemory \in \SymHeaps$.
The path constraint $\pc$ is a formula of sort $\Bool$ that accumulates assumptions along the execution path.
\end{definition}

The symbolic evaluation semantics of expressions, denoted by $\evalexprty{e}{\symlocalstore}{\tau}$, mirrors the concrete evaluation semantics (\Cref{figure:expr-semantics}).
Here, however, $\sem{\toper^\tau}$ and $\sem{\tcast^{\tau\to\tau'}}$ are \emph{function symbols} that construct SMT expressions.

\Cref{figure:symbolic-semantics} defines the symbolic operational small-step semantics of \llvm IR based on the syntax in~\Cref{figure:llvm-syntax}.
In some of the derivation rules, we use the following definitions (where $\symmo \triangleq \mobj{\bvval{\addr}{\pointerwidth}}{\bvval{\size}{\indexwidth}}{\smtarray}$):
\[
\begin{array}{ll}
\baseaddrcond(p, \symmo) & \triangleq p = \bvconst{\pointerwidth}{\addr} \\
\rangecond(p, n, \symmo) & \triangleq p < \bvadd{\pointerwidth}{p}{n} \land p \geq \bvconst{\pointerwidth}{\addr} \land \bvadd{\pointerwidth}{p}{n} < \bvadd{\pointerwidth}{\bvconst{\pointerwidth}{\addr}}{\bvconst{\pointerwidth}{\size}}
\end{array}
\]
Intuitively,
$\baseaddrcond$ defines an SMT formula asserting that $p$ points to the base address of $\symmo$,
and $\rangecond$ defines an SMT formula asserting that the interval $[p, p + n)$ is fully contained within the address interval of $\symmo$.
In the following, we discuss the derivation rules in more detail.

\begin{figure}[t]
\[
\begin{array}{c}
\ruledefsmall{Malloc}{
  \instrat{\location} = \codequote{x~\texttt{=}~\tcall^{\pointertypeof{\bvsort{8}}}~\intrinsicfunc{malloc}(\bvsort{64} ~ e)} \quad
  \evalexprty{e}{\symlocalstore}{\bvsort{64}} = n \quad
  m \models \pc \quad
  \symallocate(\symmemory, \bvextract{\indexwidth}{m[n]}) = (\symmemory', p) \quad
}
{
  \symstate{\location}{\symlocalstore}{\symstack}{\symmemory}{\pc}
  \symstep
  \symstate{\nextinstr{\location}}{\symlocalstore[x \mapsto p]}{\symstack}{\symmemory'}{\pc}
}
\\[1em]
\ruledefsmall{Free}{
  \instrat{\location} = \codequote{\tcall~\intrinsicfunc{free}(\pointertypeof{\bvsort{8}} ~ e)} \quad
  \evalexprty{e}{\symlocalstore}{\pointertypeof{\bvsort{8}}} = p \quad
  m \models \pc \quad
  \symmo \in \symmemory \quad
  \symisbaseaddr{p}{mo}{m} \quad
  \deallocate(\symmemory, \mo) = \symmemory'
}
{
  \symstate{\location}{\symlocalstore}{\symstack}{\symmemory}{\pc}
  \symstep
  \symstate{\nextinstr{\location}}{\symlocalstore}{\symstack}{\symmemory'}{\pc \land \baseaddrcond(p, \symmo)}
}
\\[1em]
\ruledefsmall{Load}{
  \instrat{\location} = \codequote{x~\texttt{=}~\tload^{\tau}~e} \quad
  \evalexprty{e}{\symlocalstore}{\pointertypeof{\tau}} = p \quad
  m \models \pc \quad
  \symmo \in \symmemory \quad
  \symisinbounds{p}{\tau}{mo}{m}
}
{
  \symstate{\location}{\symlocalstore}{\symstack}{\symmemory}{\pc}
  \symstep
  \symstate{\nextinstr{\location}}{\symlocalstore[x \mapsto \symmoread{\tau}{\symmo}{p}]}{\symstack}{\symmemory}{\pc \land \rangecond(p, |\tau|, \symmo)}
}
\\[1em]
\ruledefsmall{Store}{
  \instrat{\location} = \codequote{\tstore^{\tau}~e_1~e_2} \quad
  \evalexprty{e_1}{\symlocalstore}{\tau} = v \quad
  \evalexprty{e_2}{\symlocalstore}{\pointertypeof{\tau}} = p \quad
  m \models \pc \quad
  \symmo \in \symmemory \quad
  \symisinbounds{p}{\tau}{mo}{m}
}
{
  \symstate{\location}{\symlocalstore}{\symstack}{\symmemory}{\pc}
  \symstep
  \symstate{\nextinstr{\location}}{\symlocalstore}{\symstack}{\symheapreplace{\symmemory}{\symmo}{\symmowrite{\tau}{\symmo}{p}{v}}}{\pc \land \rangecond(p, |\tau|, \symmo)}
}
\\[1em]
\ruledefsmall{GEP}{
  \instrat{\location} = \codequote{x~\texttt{=}~\tgep~\pointertypeof{\tau}~e~\overline{(\tau_i, e_i)}} \quad
  \evalexprty{e}{\symlocalstore}{\pointertypeof{\tau}} = p \quad
  \evalexprty{e_i}{\symlocalstore}{\tau_i} = t_i
}
{
  \symstate{\location}{\symlocalstore}{\symstack}{\symmemory}{\pc}
  \symstep
  \symstate{\nextinstr{\location}}{\symlocalstore[x \mapsto \symevalgep(\tau, p, \overline{t_i})]}{\symstack}{\symmemory}{\pc}
}
\end{array}
\]
\caption{Symbolic operational small-step semantics of \llvm IR.}
\label{figure:symbolic-semantics}
\end{figure}

% TODO: mention the defined operation functions

\textbf{\textit{Malloc.}}
According to the symbolic memory model (\Cref{sec:symbolic-memory-modeling}),
the size of a symbolic memory object must be concrete.
However, as the allocation size $e$ can be evaluated into a symbolic expression,
we concretize it using an SMT model that satisfies the path constraints $\pc$.
Then, if the $\symallocate$ function (\Cref{sec:symbolic-memory-modeling}) succeeds,
we update the local store and the heap accordingly.
Otherwise, the result is undefined behavior.
Due to the concretization, this rule may produce multiple derivations.

\textbf{\textit{Free.}}
When deallocating a pointer $p$, which is a symbolic expression,
we check if there exists an SMT model $m$ satisfying $\pc$ and a memory object $\symmo$ in the heap $\symmemory$ such that $m[p]$ points to the base address of $\symmo$.
If that is the case, we remove $\symmo$ from $\symmemory$ using the $\symdeallocate$ function (\Cref{sec:symbolic-memory-modeling}) and extend $\pc$ with the condition $\baseaddrcond(p, \symmo)$.
Otherwise, the result is undefined behavior.
This rule might have multiple derivations, since $p$ may point to different memory objects depending on the SMT model.

\textbf{\textit{Load and Store.}}
% TODO: mentions the assumptions about n?
When accessing $n$ bytes from a pointer $p$,
we check if there exists an SMT model $m$ satisfying $\pc$ and a memory object $\symmo$ in $\symmemory$ such that the first $n$ addresses starting from $m[p]$ are within the address interval of $\symmo$.
If that is not the case, the result is undefined behavior.
Otherwise, we perform a read or write operation.
For a read operation,
we call the $\symmoreadfunc{\tau}$ function (\Cref{sec:symbolic-memory-modeling}) to read the value from $\symmo$ and update the local store accordingly.
For a write operation,
we call the $\symmowritefunc{\tau}$ function (\Cref{sec:symbolic-memory-modeling}) to update the contents of $\symmo$ accordingly.

\textbf{\textit{GetElementPtr.}}
The symbolic semantics of $\tgep$ mirrors the concrete semantics,
while using SMT expressions for offset and address calculations.
The function $\symevalgep$ in~\Cref{alg:symbolic-gep} defines a simplified symbolic semantics of the $\tgep$ instruction.
While the concrete version may return $\tundef$ or $\tpoison$,
the symbolic version always returns an SMT expression.

\begin{algorithm}[t]
\caption{Simplified symbolic semantics of $\tgep$ (without attributes).}
\label{alg:symbolic-gep}
\begin{algorithmic}[1]

\Function{compute-offset}{$\tau, o, \overline{j}$}
    \If {$|\overline{j}| = 0$}
        \State \Return $o$
    \EndIf
    \State $j_1 :: \overline{j'} \gets \overline{j}$
    \If {$\tau = \arraytype{\tau'}{n}$}
        \State $o \gets \bvadd{\bvsort{\indexwidth}}{o}{\bvmul{\bvsort{\indexwidth}}{\bvsext{\indexwidth}{j_1}}{\bvconst{\indexwidth}{|\tau'|}}}$ \label{line:array-index}
        \State \Return \Call{compute-offset}{$\tau', o, \overline{j'}$}
    \EndIf
\EndFunction

\Function{eval-gep}{$\tau, p, \overline{j}$}
    \State $j_1 :: \overline{j'} \gets \overline{j}$
    \State $o \gets \bvmul{\bvsort{\indexwidth}}{\bvsext{\indexwidth}{j_1}}{\bvconst{\indexwidth}{|\tau|}}$ \label{line:first-index}
    \State $r \gets \Call{compute-offset}{\tau, o, \overline{j'}}$
    \State \Return $\bvadd{\bvsort{\pointerwidth}}{p}{\bvzext{\pointerwidth}{r}}$
\EndFunction

\end{algorithmic}
\end{algorithm}

Note that our symbolic semantics does not model the values $\tundef$ and $\tpoison$.
While this might initially appear to be a limitation,
modeling these values in the symbolic semantics is not mandatory.
The design of the symbolic semantics is a choice of abstraction level,
and we intentionally omit these values to simplify SMT reasoning.
Indeed, as will be seen in~\Cref{sec:program-safety},
our design choice allows to establish the desired properties with respect to the concrete semantics.

In the following, we define when a symbolic state is considered to be an error state.
To do so, we use the following definitions:
\[
\begin{array}{ll}
    \syminvalidpointer(\symmemory, \pc, p, n) & \triangleq \exists m. \ \forall \symmo \in \symmemory. \ m \not \models (\pc \land \rangecond(p, n, \symmo)) \\
    \syminvalidbaseaddress(\symmemory, \pc, p) & \triangleq \exists m. \ \forall \symmo \in \symmemory. \ m \not \models (\pc \land \baseaddrcond(p, \symmo))
\end{array}
\]
Intuitively,
$\syminvalidpointer$ states that there is an SMT model $m$ satisfying $\varphi$ under which $p$ does not belong to any address interval,
and $\syminvalidbaseaddress$ states that under such a model, $p$ does not point to any base address.

\begin{definition}
\label{def:sym-error-state}
A symbolic state $s \triangleq \symstate{\location}{\symlocalstore}{\symstack}{\symmemory}{\pc}$ is an error state, denoted by $\errorsymstate(s)$,
if one of the following holds:
\begin{enumerate}[left=0pt]
    \item $\instrat{\location} = \textup{$\codequote{\tunreachable}$}$
    \item $\instrat{\location} = \textup{$\codequote{x \ \texttt{=} \ op \ \bvsort{w} \ e_1 \ e_2}$}$ and one of the following holds (where $t_j \triangleq \evalexprty{e_j}{\symlocalstore}{\bvsort{w}}$):
    \begin{enumerate}[left=-3pt]
        \item \textup{$op \in \{\opcode{udiv}, \opcode{sdiv}, \opcode{urem}, \opcode{srem}\}$} and $\sat{\pc \land (t_2 = \bvconst{w}{0})}$
        \item \textup{$op = \opcode{sdiv}$} and $\sat{\pc \land (t_1 = \bvconst{w}{(-2^{w - 1})}) \land (t_2 = \bvconst{w}{(-1)})}$
        \item \textup{$op \in \{\opcode{shl}, \opcode{lshr}, \opcode{ashr}\}$} and $\sat{\pc \land (t_2 \geq w)}$
    \end{enumerate}
    \item $\instrat{\location} = \textup{$\codequote{x \ \texttt{=} \ \tcall^{\pointertypeof{\bvsort{8}}} \ \intrinsicfunc{malloc}(\bvsort{64} \ e)}$}$ and
        \textup{$\sat{\pc \land (t > \bvconst{64}{(2^{\indexwidth - 1} - 1)}}$} (where $t \triangleq \evalexprty{e}{\symlocalstore}{\tau}$).
    \item $\instrat{\location} = \textup{$\codequote{\tcall~\intrinsicfunc{free}(\pointertypeof{\bvsort{8}} e)}$}$, \textup{$\evalexprty{e}{\symlocalstore}{\pointertypeof{\bvsort{8}}} = p$}, and $\syminvalidbaseaddress(\symmemory, \pc, p)$.
    \item $\instrat{\location} = \textup{$\codequote{x \ \texttt{=} \ \tload^{\tau} \ e}$}$, \textup{$\evalexprty{e}{\symlocalstore}{\pointertypeof{\tau}} = p$}, and $\syminvalidpointer(\symmemory, \pc, p, |\tau|)$.
    \item $\instrat{\location} = \textup{$\codequote{\tstore^{\tau} \ e_1 \ e_2}$}$, \textup{$\evalexprty{e_2}{\symlocalstore}{\pointertypeof{\tau}} = p$}, and $\syminvalidpointer(\symmemory, \pc, p, |\tau|)$.
    %\item $\instrat{\location} = \textup{$\codequote{x \ \texttt{=} \ \tgep~\pointertypeof{\tau} \ e \ \overline{(\tau_i, e_i)}}$}$ and $\sat{\varphi \land \overflowcond(\tau, \overline{k_i})}$
    %      (where $k_i \triangleq \evalexprty{e_i}{\localstore}{\tau_i}$).
    \item $\instrat{\location} = \textup{$\codequote{x \ \texttt{=} \ \tgep~\pointertypeof{\tau} \ e \ \overline{(\tau_i, e_i)}}$}$ and there exists an SMT model $m$ such that:
    \begin{enumerate}
        \item $m \models \varphi$
        \item $\evalgep(\tau, p, \overline{v_i}) = \tpoison$ (where $p \triangleq m[\evalexprty{e}{\localstore}{\pointertypeof{\tau}}]$ and $v_i \triangleq m[\evalexprty{e_i}{\localstore}{\tau_i}]$)
    \end{enumerate}
\end{enumerate}
%where $\overflowcond(\tau, \overline{k_i})$ constructs a condition that encodes the fact that there are no overlfows when evaluating the $\tgep$ instruction,
\end{definition}
